Left-linear rules in adhesive categories

### 2024-02-13 icalp/fusione.tex
poco

There is, however, a further problem. In the literature on linear
rewriting systems, a central result is the a characterisation of
switch equivalence of sequences of steps based on what are called processes or
consistent permutations~\cite{Handbook, tobias}.\todo{Handbook, graph processes, tobias}
In turn this is the key
to derive some simple but fundamental properties for linear rewriting
systems, namely

\begin{enumerate}
\item \emph{consistency of switching}: when transforming
  sequences of rewriting steps into equivalent ones by switching independent steps, the
  result not depends on the order in which switchings are applied, but
  only on the resulting permutation;
\item \emph{globality of independence}: if two independent steps are
  moved forward or backward due to the application of switchings to a
  sequence, they remain independent.
\end{enumerate}
Properties (1) and (2) allow one to deduce a canonical form for equivalence proofs 
between sequences of rewriting steps seen as concurrent computations.

Unfortunately, for left-linear rewriting systems the characterisation
of switch equivalence based on consistent permutations fails: the
existence of a consistent permutation is only necessary but not
sufficient, hence it cannot be used for proving (1) and (2).
%
This leads to asking whether the properties above hold.
We show that actually much of the theory can be retained: consistency
of switching still holds, while globality of independence holds in a
weak form, conceptually linked to the fact that fusions introduce a
form of disjunctive causality. Finally, we prove that a canonical form
for switching sequences can also be recovered.

@@

$\mathcal{M}$-adhesive categories are the right context in which to
perform abstract rewriting using what is called the ``double pushout
approach'' (DPO). We will recall the basic definitions and properties
of this approach to abstract rewriting.

@@

Intuitively, the existence of the independence pair
captures the possibility of switching the application of the two rules:
$f_0 \circ i_1$ is a match for $\rho_1$ in $G_0$ and the
existence of $i_0 : R_0 \to D_1$ indicates that $\rho_1$ does not delete
items in the image of $K_0$ and thus $\rho_0$ can be applied after
$\rho_1$.

@@

As already mentioned, for linear rewriting systems, when viewing
derivations as concurrent computations, it is canonical to identify
derivations which are equal ``up to switching'', i.e., which differs
only for the order of indepedent steps. The same notion can be given
for left-linear rewriting systems by relying on notion of switch.

@@

\begin{definition}[Switch equivalence]
  \label{de:switch-equivalence}
  Let $(\X, \R)$ be a left-linear DPO rewriting system.  Let also
  $\der{D}, \der{E} : G \Mapsto H$ be derivations with the same
  length, $\der{D}=\{\dder{D}_{i}\}_{i=0}^n$ and
  $\der{E}=\{\dder{E}_{i}\}_{i=0}^n$. If
  $\dder{D}_i \cdot \dder{D}_{i+1}$ is a switch of
  $\dder{E}_i \cdot \dder{E}_{i+1}$ for some $i \in [0,n-1]$ we write
  $\der{D} \shift{\transp{i}} \der{E}$.

  A \emph{switching sequence} is  a sequence $\{\der{D}_{k}\}_{i=0}^m$ of derivations such that 
  \[\der{D}_0 \shift{\transp{i_1}} \der{D}_1 \shift{\transp{i_2}} \ldots \shift{\transp{i_m}}
  \der{D}_m\]
 In such case, we will often denote by $\nu_{k}$ the transposition $\transp{i_k}$. Given $k \leq h$,
 we denote by $\nu_{k,h}$ the composition
 $\nu_h \circ \nu_{h-1} \circ \ldots \nu_k$. We will say that the
 switching sequence consists of inversions if for all
 $k \in \interval[0]{m}$ the transposition $\nu_k$ is an inversion for
 $\nu_{1,m}$.

@@

\begin{example}
  \label{ex:seq-ind}
  Consider a rewriting system in $\cat{Graph}$, the category of
  directed graphs and graph morphisms, which is adhesive. The set of
  rules $\R = \{ \rho_0, \rho_1, \rho_2\}$ consists of three rules as
  depicted in Fig.~\ref{fi:rules}, where numbers are used to represent
  the mappings from the interface to the left- and right-hand
  sides. Rules $\rho_0$ and $\rho_1$ are linear: $\rho_0$ generates a
  new node, while $\rho_1$ deletes an edge. Instead, $\rho_2$ is non
  linear as it ``merges'' two nodes.

@@

\begin{example}
  \label{ex:diff2}
  In light of \Cref{thm:church}, \Cref{ex:diff1} provides an example
  of an independence pair which is not strong, and thus of a non strong enforcing
  adhesive rewriting system. Instead the category $\cat{Graph}$ is
  strong enforcing by the result in .... 
\end{example}

@@

\begin{definition} Let $\X$ be an $\mathcal{M}$-adhesive category, we say that the pair $(\X, \mathcal{M})$ is of \emph{type $\mathbb{B}^+$} if (at least) one of the following holds true:
	\begin{enumerate}
		\item $(\X, \mathcal{M})$ is of type $\mathbb{B}$;
	\item $\X$ is $\Y/Y$ and $\mathcal{M}=\mathcal{N}/Y$for some $(\Y, \mathcal{N})$ of type $\mathbb{B}^+$ and $Y\in \Y$;
 	\item $\X$ is $Y/\Y$ and $\mathcal{M}=Y/\mathcal{N}$for some $(\Y, \mathcal{N})$ of type $\mathbb{B}^+$ and $Y\in \Y$:
 	\item $\X$ is $\Y^{\A}$ and $\mathcal{M}=\mathcal{N}^{\mathcal{A}}$ for some category category $\A$ and $(\Y, \mathcal{N})$ of type $\mathbb{B}^+$.
	\end{enumerate}
\end{definition}
\begin{example}
	The pair $(\mathbf{Graph}, \mon(\mathbf{Graph}))$ made by the topos of graphs and its class of mons is of type $\mathbb{B}^+$  but not of type $\mathbb{B}$. Indeed not every monomorphism of graphs is a coproduct coprojection.
\end{example}

@@

Using \Cref{thm:good1} we finally get the main result of our appendix.
\begin{corollary}
Let $(\X, \mathcal{M})$ be a pair of type $\mathbb{B}^+$, then every left-linear DPO-rewriting system is strong enforcing.
\end{corollary}

@@

\begin{lemma}\label{lem:idevero}
	Let $\der{D}=\{\dder{D}_{i}\}_{i=0}^n$ and $\der{D}'=\{\dder{D}'_i\}_{i=0}^n$ be two non-empty derivations in a linear DPO-rewriting system $(\X, \R)$ . Let  $\sigma:[0,n-1]\to [0,n-1]$  is a consistent permutation and suppose that the set 
	\[E_{\sigma}=\{i\in [0,n-1]\mid \sigma(i)=i\}\]
	is non empty with maximum $s_\sigma$. Then $\{\dder{D}_i\}_{i=0}^{s_\sigma}$ and $\{\der{D}'_i\}_{i=0}^{s_\sigma}$ are abstraction equivalent.  
\end{lemma}

>>>>

There is, however, a further problem. In the literature on linear
rewriting systems, a central result is the characterisation of
switch equivalence of sequences of steps based on what are called processes or
consistent permutations~\cite{Handbook, BaldanCHKS06, tobias}.\todo{Handbook, graph processes, tobias}
In turn this is the key
to derive some simple but fundamental properties for linear rewriting
systems, namely

\begin{enumerate}
\item \emph{consistency of switching}: when transforming
  sequences of rewriting steps into equivalent ones by switching independent steps, the
  result does not depend on the order in which switchings are applied, but
  only on the resulting permutation;
\item \emph{globality of independence}: if two independent steps are
  moved forward or backward due to the application of switchings to a
  sequence, they remain independent.
\end{enumerate}
Properties (1) and (2) allow one to deduce a canonical form for equivalence proofs 
between sequences of rewriting steps seen as concurrent computations.

Unfortunately, for left-linear rewriting systems the characterisation
of switch equivalence based on consistent permutations fails: the
existence of a consistent permutation is only necessary but not
sufficient, hence it cannot be used for proving (1) and (2).
%
This leads to asking whether the properties above hold.
We show that actually much of the theory can be retained: consistency
of switching still holds, while globality of independence holds in a
weaker form, conceptually linked to the fact that fusions introduce a
form of disjunctive causality. Finally, we prove that a canonical form
for switching sequences can also be recovered.

@@

$\mathcal{M}$-adhesive categories are the right context in which to
perform abstract rewriting using what is called the ``Double-Pushout
approach'' (DPO). We will recall the basic definitions and properties
of this approach to abstract rewriting.

@@

Intuitively, the existence of the independence pair
captures the possibility of switching the application of the two rules:
$f_0 \circ i_1$ is a match for $\rho_1$ in $G_0$ and the
existence of $i_0 : R_0 \to D_1$ means that $\rho_1$ does not delete
items in the image of $K_0$, thus $\rho_0$ can be applied after
$\rho_1$.

@@

As already mentioned, for linear rewriting systems, when viewing
derivations as concurrent computations, it is canonical to identify
derivations which are equal ``up to switching'', i.e., which differs
only for the order of indepedent steps. The same notion can be given
for left-linear rewriting systems by relying on the notion of switch.

@@

\begin{definition}[Switch equivalence]
  \label{de:switch-equivalence}
  Let $(\X, \R)$ be a left-linear DPO rewriting system.  Let also
  $\der{D}, \der{E} : G \Mapsto H$ be derivations with the same
  length, $\der{D}=\{\dder{D}_{i}\}_{i=0}^n$ and
  $\der{E}=\{\dder{E}_{i}\}_{i=0}^n$. If
  $\dder{D}_i \cdot \dder{D}_{i+1}$ is a switch of
  $\dder{E}_i \cdot \dder{E}_{i+1}$ for some $i \in [0,n-1]$ \todo{A: ma manca ``and  $\dder{D}_j = \dder{E}_j$ for each $j \not \in \{i,i+1\}$''??} we write
  $\der{D} \shift{\transp{i}} \der{E}$.

  A \emph{switching sequence} is  a sequence $\{\der{D}_{k}\}_{k=0}^m$ of derivations such that 
  \[\der{D}_0 \shift{\transp{i_1}} \der{D}_1 \shift{\transp{i_2}} \ldots \shift{\transp{i_m}}
  \der{D}_m\]
 In such case, we will\todo{A: io cambierei tutti i ``we will'' in ``we''.} often denote by $\nu_{k}$ the transposition $\transp{i_k}$. Given $k \leq h$,
 we denote by $\nu_{k,h}$ the composition
 $\nu_h \circ \nu_{h-1} \circ \ldots \nu_k$. We will say that the
 switching sequence consists of inversions if for all
 $k \in \interval[0]{m}$ the transposition $\nu_k$ is an inversion for
 $\nu_{1,m}$.

@@

\begin{example}
  \label{ex:seq-ind}
  Consider a rewriting system in $\cat{Graph}$, the category of
  directed graphs and graph morphisms, which is adhesive. The set of
  rules $\R = \{ \rho_0, \rho_1, \rho_2\}$ consists of three rules as
  depicted in Fig.~\ref{fi:rules}, where numbers are used to represent
  the mappings from the interface to the left- and right-hand
  sides. Rules $\rho_0$ and $\rho_1$ are linear: $\rho_0$ generates a
  new node, while $\rho_1$ deletes an edge. Instead, $\rho_2$ is not
  linear as it ``merges'' two nodes.

@@

\begin{example}
  \label{ex:diff2}
  In light of \Cref{thm:church}, \Cref{ex:diff1} provides an example
  of an independence pair which is not strong, and thus of a non-strong enforcing
  adhesive rewriting system. Instead every left-linear system in category $\cat{Graph}$ is
  strong enforcing, by the results of Appendix~\ref{app:fill}, summarized in~\Cref{ex:graph-type-bb} and~\Cref{cor:llr-is-strEnf}.
\end{example}

@@

\begin{definition} Let $\X$ be an $\mathcal{M}$-adhesive category, we say that the pair $(\X, \mathcal{M})$ is of \emph{type $\mathbb{B}^+$} if (at least) one of the following holds true:
	\begin{enumerate}
		\item $(\X, \mathcal{M})$ is of type $\mathbb{B}$;
	\item $\X$ is $\Y/Y$ and $\mathcal{M}=\mathcal{N}/Y$for some $(\Y, \mathcal{N})$ of type $\mathbb{B}^+$ and $Y\in \Y$;
 	\item $\X$ is $Y/\Y$ and $\mathcal{M}=Y/\mathcal{N}$for some $(\Y, \mathcal{N})$ of type $\mathbb{B}^+$ and $Y\in \Y$:
 	\item $\X$ is $\Y^{\A}$ and $\mathcal{M}=\mathcal{N}^{\mathcal{A}}$ for some category category $\A$ and $(\Y, \mathcal{N})$ of type $\mathbb{B}^+$.
	\end{enumerate}
\end{definition}
\begin{example}
  \label{ex:graph-type-bb}
	The pair $(\mathbf{Graph}, \mon(\mathbf{Graph}))$ made by the topos of graphs and its class of mons is of type $\mathbb{B}^+$  but not of type $\mathbb{B}$. Indeed not every monomorphism of graphs is a coproduct coprojection.
\end{example}

@@

Using \Cref{thm:good1} we finally get the main result of our appendix.
\begin{corollary}
  \label{cor:llr-is-strEnf}
Let $(\X, \mathcal{M})$ be a pair of type $\mathbb{B}^+$, then every left-linear DPO-rewriting system is strong enforcing.
\end{corollary}

@@

\begin{lemma}\label{lem:idevero}
	Let $\der{D}=\{\dder{D}_{i}\}_{i=0}^n$ and $\der{D}'=\{\dder{D}'_i\}_{i=0}^n$ be two non-empty derivations in a linear DPO-rewriting system $(\X, \R)$ . Let  $\sigma:[0,n-1]\to [0,n-1]$  is a consistent permutation and suppose that the set 
	\[E_{\sigma}=\{i\in [0,n-1]\mid \sigma(i)=i\}\]
	is not empty with maximum $s_\sigma$. Then $\{\dder{D}_i\}_{i=0}^{s_\sigma}$ and $\{\der{D}'_i\}_{i=0}^{s_\sigma}$ are abstraction equivalent.  
\end{lemma}
